---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

# datavolley

[![Travis-CI Build Status](https://travis-ci.org/raymondben/datavolley.svg?branch=master)](https://travis-ci.org/raymondben/datavolley)

An R package for reading DataVolley scouting files.

See also this [DataVolley file validator](https://untangl.shinyapps.io/dvalidate/), which is built on the datavolley package.

## Installation

```{r eval=FALSE}
library(devtools)
install_github("raymondben/datavolley")
```

## Example

Read one of the example data files bundled with the package:
```{r}
library(datavolley)
x <- read_dv(dv_example_file(), insert_technical_timeouts=FALSE)
summary(x)
```

Number of serves by team:

```{r}
serve_idx <- find_serves(plays(x))
table(plays(x)$team[serve_idx])
```

Distribution of serve run lengths:

```{r}
serve_run_info <- find_runs(plays(x)[serve_idx,])
table(unique(serve_run_info[,c("run_id","run_length")])$run_length)
```

Heatmap of attack rate by court position:

```{r}
library(ggplot2)
library(dplyr)
## calculate attack frequency by zone, per team
attack_rate <- plays(x) %>% dplyr::filter(skill=="Attack") %>%
    group_by(team, start_zone) %>% dplyr::summarize(n_attacks=n()) %>%
    mutate(rate=n_attacks/sum(n_attacks)) %>% ungroup

## add x,y coordinates associated with the zones
attack_rate <- cbind(attack_rate,ggxy(attack_rate$start_zone,end="lower"))

## for team 2, these need to be on the top half of the diagram
tm2 <- attack_rate$team==teams(x)[2]
attack_rate[tm2, c("x", "y")] <- ggxy(attack_rate$start_zone, end="upper")[tm2, ]

ggplot(attack_rate, aes(x, y, fill=rate)) + geom_tile() + ggcourt(labels=teams(x)) +
    scale_fill_gradient2(name="Attack rate")
```

Or using arrows to show the starting and ending zones of attacks:

```{r}
## first tabulate attacks by starting and ending zone
attack_rate <- plays(x) %>% dplyr::filter(team==teams(x)[1] & skill=="Attack") %>%
  group_by(start_zone, end_zone) %>% tally() %>% ungroup

## convert counts to rates
attack_rate$rate <- attack_rate$n/sum(attack_rate$n)

## discard zones with zero attacks or missing location information
attack_rate <- attack_rate %>% dplyr::filter(rate>0 & !is.na(start_zone) & !is.na(end_zone))

## add starting x,y coordinates
attack_rate <- cbind(attack_rate, ggxy(attack_rate$start_zone, end="lower", xynames=c("sx","sy")))

## and ending x,y coordinates
attack_rate <- cbind(attack_rate, ggxy(attack_rate$end_zone, end="upper", xynames=c("ex","ey")))

## plot in reverse order so largest arrows are on the bottom
attack_rate <- attack_rate %>% dplyr::arrange(desc(rate))

p <- ggplot(attack_rate,aes(x,y,col=rate)) + ggcourt(labels=c(teams(x)[1],""))
for (n in 1:nrow(attack_rate))
    p <- p + geom_path(data=data.frame(x=c(attack_rate$sx[n], attack_rate$ex[n]),
                                       y=c(attack_rate$sy[n],attack_rate$ey[n]),
                                       rate=attack_rate$rate[n]),
        aes(size=rate), lineend="round", arrow=arrow(ends="last", type="closed"))
p + scale_fill_gradient(name="Attack rate") + guides(size="none")
```

## Troubleshooting

If you see unexpected behaviour, try `read_dv(...,do_warn=TRUE)` to obtain more diagnostic information during the process of reading and parsing the DataVolley file. Also check the text encoding specified to `read_dv` (did you specify one??)


## More

For more information about DataVolley, see http://www.dataproject.com/IT/en/Volleyball.

